Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-21 paper.tex
changes from Yan
Fix line break (again)

\subsection{Data preprocessing}
\subsubsection{Sequence-level preprocessing}
The findings of this study are based on sequences and metadata available on GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
% FIXME
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed if they (1) had ambiguous collection dates; (2) were collected before January 01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host. Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag 2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG inference, sequences were excluded if they had a ``bad'' quality control status in any of the four Nextclade columns (``qc.missingData.status'', ``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status''). These excluded genomes are enriched for low-coverage, contaminated, or misassembled sequences.

@@

\subsubsection{Exclusion of ``time travellers''}
Including samples with incorrectly reported collection dates (e.g., an Alpha sample purportedly collected in 2020 from the United States, before Alpha appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce artefacts that distorted the topology of ARGs inferred early on in the study. Hence, to exclude such potential ``time travellers'' we employed two filters. The first was to remove any sequence with a collection date that pre-dated the time to the most recent common ancestor (tMRCA) of its corresponding clade. The tMRCA for each clade was obtained from a Nextstrain GISAID reference tree (downloaded on August 22, 2022) where we used the lower bound of the 95\% confidence interval of each clade as the minimum date cut-off. The second filter was a simple threshold on the time delay between the collection date and submission date (maximum submission delay, or MSD). Indeed, we found that ARGs inferred from the MSD-filtered genome set had noticeably fewer branches with a high number of reversions. After some preliminary analysis, a MSD threshold value of 30 days was used when building the ARGs described herein. Future work could estimate collection dates by adding these potential time travellers back on the tree, allowing an automated assessment of collection date discrepancies.

>>>>

\subsection{Data preprocessing}

\subsubsection{Sequence-level preprocessing}
The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
% FIXME
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
if they (1) had ambiguous collection dates; (2) were collected before January
01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, sequences were excluded if they had a ``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and
`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
contaminated, or misassembled sequences.

@@

\subsubsection{Exclusion of ``time travellers''}

Including samples with incorrectly reported collection dates (e.g., an Alpha
sample purportedly collected in 2020 from the United States, before Alpha
appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
artefacts that distorted the topology of ARGs inferred early on in the study.
Hence, to exclude such potential ``time travellers'' we employed two filters.
The first was to remove any sequence with a collection date that pre-dated the
time to the most recent common ancestor (tMRCA) of its corresponding clade. The
tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
(downloaded on August 22, 2022) where we used the lower bound of the 95\%
confidence interval of each clade as the minimum date cut-off. The second
filter was a simple threshold on the time delay between the collection date and
submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
inferred from the MSD-filtered genome set had noticeably fewer branches with a
high number of reversions. After some preliminary analysis, a MSD threshold
value of 30 days was used when building the ARGs described herein. Future work
could estimate collection dates by adding these potential time travellers back
on the tree, allowing an automated assessment of collection date discrepancies.
